Non-Vertical Cultural Transmission, Assortment, \\and the Evolution of Cooperation

### 2020-09-14 ms/main.tex
revise result 3

Assume $T_A<T_B$.
For $\beta_3<0$ we need $\hat\gamma_1<c$.
For $\beta_3<\beta_1$ we need $c\big[v(1-T_B) + (1-v)(T_A-T_B)\big] > bv\alpha T_B + (1+b)(T_A-T_B)$.
Note that the expression in the square brackets is positive if and only if $v > \hat v$.
So, for $\beta_3<\beta_1$ we need $v > \hat v$ and $\hat\gamma_2 < c$ or $v < \hat v$ and $c < \hat\gamma_2$.
So, for $0<\beta_3<\beta_1$ we need $v > \hat v$ and $\hat\gamma_2 < c < \hat\gamma_1$, or $v < \hat v$ and $c < \min(\hat\gamma_1, \hat\gamma_2)$; and 
for $\beta_1<\beta_3$ we need $v > \hat v$ and $c<\hat\gamma_2$ or $v < \hat v$ and $\hat\gamma_2<c$.
However, some of these conditions cannot be met, since $v < \hat v$ implies $c<1<\hat\gamma_2$.
A similar analysis can be made when $T_A>T_B$.

@@

% in gamma
\begin{enumerate}
\item \emph{Fixation of cooperation}: 
	if \emph{(i)} $T_A \ge T_B$ and $c < \hat\gamma_1$; or 
	if \emph{(ii)} $T_A < T_B$ and $v>\hat v$ and $c < \hat\gamma_2$.
\item \emph{Fixation of defection}: 
    if \emph{(iii)} $T_A \ge T_B$ and $\hat\gamma_2 < c$; or 
	if \emph{(iv)} $T_A < T_B$ and $\hat\gamma_1 < c$.
\item \emph{Stable Coexistence}: 
    if \emph{(vi)} $T_A < T_B$ and $v>\hat{v}$ and $\hat\gamma_2 < c < \hat\gamma_1$. 
\item \emph{Unstable coexistence}:
    if \emph{(vii)} $T_A > T_B$ and $\hat\gamma_1 < c < \hat\gamma_2$.
\end{enumerate}

@@

%This result illustrates the crucial role of vertical and oblique transmission: the same cost of cooperation $c$ can lead to remarkably different outcomes if the vertical transmission rate $v$ is below or above the threshold $\hat v$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Discussion
\section*{Discussion}
We studied the evolution of cooperation under non-vertical transmission using deterministic discrete-time evolutionary models with fitnesses in the form of payoffs from a prisoner's dilemma game. 
Under oblique and horizontal cultural transmission, a horizontal transmission bias for the cooperative phenotype was found to be sufficient and necessary for evolution of cooperation (Result~\autoref{result:obli_hori}).
Under a combination of vertical, oblique, and horizontal transmission, cooperation or defection can either fix or coexist at a stable polymorphism, depending on the relationship between the cost and benefit of cooperation, the horizontal bias, and the association between cooperation and transmission (Results~\autoref{result:vert_hori} and~\autoref{result:vert_obli_hori}).
Remarkably, stable coexistence between cooperation and defection can be maintained if horizontal transmission is biased for defection ($T_B>T_A$) and both the cost of cooperation and social association are intermediate (yellow areas in \autoref{fig:results2}).
% TODO figure for result 3
% TODO something about v-hat

We tested the applicability of our analytic results to a complex spatial stochastic model (\autoref{fig:spatial}). Indeed, our derived condition provide a good approximation to simulation results, demonstrating that association between cooperation and transmission can arise in a structured population simply because the choice of individuals for social interactions is small.

@@

\citet{woodcock2006significance} has previously argued the significance of non-vertical transmission for the evolution of cooperation. He executed simulations with prisoner's dilemma payoffs but without horizontal transmission or social association ($\alpha=0$). Nevertheless, his results demonstrated that it is possible to sustain altruistic behavior via cultural transmission for a significant duration of time.
Our results provide strong evidence for his hypothesis that horizontal transmission could have an important role in the evolution of cooperation.

To understand the role of horizontal transmission, we first discuss the role of \emph{assortment}.
\citet{Eshel1982} showed that altruism can evolve when there is enough \emph{assortative meeting}, namely, a tendency for individuals to interact within their phenotypic group.
\citet{Fletcher2009assortment} have further argued that a general explanation for the evolution of altruism is given by \emph{assortment}: the correlation between individuals that carry an altruistic trait and the amount of altruistic behavior in their interaction group.
They therefore suggested that to explain the evolution of altruism, we should seek mechanisms that generate such assortment, like population structure, repeated interactions, and individual recognition.
Our results highlight another mechanism for generating assortment: an association between social interactions and horizontal transmission that creates a correlation between one's partner for interaction and partner for transmission.
This mechanism does not require population structure, repeated interactions, or individual recognition.
We show that high levels of such social association greatly increase the potential for evolution of cooperation (\autoref{fig:results2}).
With enough social association ($\alpha > \big(c(1-T_B) + (T_B-T_A)\big) / b T_A$), cooperation can increase in frequency when initially rare even when there is horizontal transmission bias against it ($T_B>T_A$, see Result~\autoref{result:vert_hori}).

>>>>

Now, assume $T_A<T_B$.
For $\beta_3<0$ we need $\hat\gamma_1<c$.
For $\beta_3<\beta_1$ we need $c\big[v(1-T_B) + (1-v)(T_A-T_B)\big] > bv\alpha T_B + (1+b)(T_A-T_B)$.
Note that the expression in the square brackets is positive if and only if $v > \hat v$.
So, for $\beta_3<\beta_1$ we need $v > \hat v$ and $\hat\gamma_2 < c$ or $v < \hat v$ and $c < \hat\gamma_2$.
So, for $0<\beta_3<\beta_1$ we need $v > \hat v$ and $\hat\gamma_2 < c < \hat\gamma_1$, or $v < \hat v$ and $c < \min(\hat\gamma_1, \hat\gamma_2)$; and 
for $\beta_1<\beta_3$ we need $v > \hat v$ and $c<\hat\gamma_2$ or $v < \hat v$ and $\hat\gamma_2<c$.
However, some of these conditions cannot be met, since $v < \hat v$ implies $c<1<\hat\gamma_2$.
A similar analysis can be made when $T_A>T_B$.

@@

% in gamma
\begin{enumerate}
\item \emph{Fixation of cooperation}: 
	if \emph{(i)} $T_A \ge T_B$ and $c < \hat\gamma_1$; or 
	if \emph{(ii)} $T_A < T_B$ and $v>\hat v$ and $c < \hat\gamma_2$.
\item \emph{Fixation of defection}: 
    if \emph{(iii)} $T_A \ge T_B$ and $\hat\gamma_2 < c$; or 
	if \emph{(iv)} $T_A < T_B$ and $\hat\gamma_1 < c$.
\item \emph{Stable Coexistence}: 
    if \emph{(v)} $T_A < T_B$ and $v>\hat{v}$ and $\hat\gamma_2 < c < \hat\gamma_1$. 
\item \emph{Unstable coexistence}:
    if \emph{(vi)} $T_A > T_B$ and $\hat\gamma_1 < c < \hat\gamma_2$.
\end{enumerate}

@@

\emph{//additional corollaries//}

%Note that $\sigma \frac{\partial \gamma_1}{\partial v} = \sigma (T_B-T_A)$ and $\sigma \frac{\partial \hat\gamma_2}{\partial v} = \sigma (T_A-T_B)(b \alpha T_B - (1+b)(1-T_A))$

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Discussion
\section*{Discussion}
We studied the evolution of cooperation under non-vertical transmission using deterministic discrete-time evolutionary models with fitnesses in the form of payoffs from a prisoner's dilemma game. 
Under oblique and horizontal cultural transmission, a horizontal transmission bias for the cooperative phenotype was found to be sufficient and necessary for evolution of cooperation (Result~\autoref{result:obli_hori}).
Under a combination of vertical, oblique, and horizontal transmission, cooperation or defection can either fix or coexist at a stable polymorphism, depending on the relationship between the cost and benefit of cooperation, the horizontal bias, and the association between cooperation and transmission (Results~\autoref{result:vert_hori} and~\autoref{result:vert_obli_hori}).
Remarkably, stable coexistence between cooperation and defection can be maintained if horizontal transmission is biased for defection ($T_A<T_B$) and both the cost of cooperation and social association are intermediate (yellow areas in \autoref{fig:results2}).
% TODO figure for result 3
% TODO something about v-hat

We tested the applicability of our analytic results to a complex spatial stochastic model. Our derived condition provide a good approximation to simulation results  (\autoref{fig:spatial}), demonstrating that association between cooperation and transmission can arise in a structured population simply because the choice of individuals for social interactions is small.

@@

\citet{woodcock2006significance} has previously argued the significance of non-vertical transmission for the evolution of cooperation. He executed simulations with prisoner's dilemma payoffs but without horizontal transmission or social association ($\alpha=0$). Nevertheless, his results demonstrated that it is possible to sustain altruistic behavior via cultural transmission for a significant duration of time.
Our results provide strong evidence for his hypothesis that horizontal transmission can have an important role in the evolution of cooperation.

To understand the role of horizontal transmission, we first discuss the role of \emph{assortment}.
\citet{Eshel1982} showed that altruism can evolve when there is enough \emph{assortative meeting}, namely, a tendency for individuals to interact within their phenotypic group.
\citet{Fletcher2009assortment} have further argued that a general explanation for the evolution of altruism is given by \emph{assortment}: the correlation between individuals that carry an altruistic trait and the amount of altruistic behavior in their interaction group; see also \citet{Bijma2010assortment}.
They therefore suggested that to explain the evolution of altruism, we should seek mechanisms that generate such assortment, like population structure, repeated interactions, and individual recognition.
Our results highlight another mechanism for generating assortment: an association between social interactions and horizontal transmission that creates a correlation between one's partner for interaction and partner for transmission.
This mechanism does not require population structure, repeated interactions, or individual recognition.
We show that high levels of such social association greatly increase the potential for evolution of cooperation (\autoref{fig:results2}).
With enough social association ($\alpha > \big(c(1-T_B) + (T_B-T_A)\big) / b T_A$), cooperation can increase in frequency when initially rare even when there is horizontal transmission bias against it ($T_A<T_B$, see Result~\autoref{result:vert_hori}).
